### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/1-introduction.tex
Updated paper to responde to coauthor comments. Will use the updated diff with master to finish revising 'response.tex' in upcoming commits.

The code by Fritsch [1982] for $C^1$ cubic spline interpolation is the
predominantly utilized code for constructing monotone interpolants at
present. Theory has been provided for the quintic case [Ulrich and
Watson 1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software
exists. This work improves upon the algorithm presented by Lux et al.
[2020] by refactoring computations for improved numerical stability,
estimating minimum curvature derivatives at breakpoints with a
quadratic facet model, and using a binary search to reduce the
magnitude of the modifications made to initial derivative estimates
when constructing a monotone spline interpolant.

The importance of piecewise quintic interpolation over lower order
approximations can be simply observed. In general, the order of a
polynomial determines the number of function (and derivative) values
it can interpolate, which in turn determines the growth rate of error
away from interpolated values. $C^2$ quintic (order six) splines
match the function value and two given derivatives at each breakpoint.
This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
necessary and sufficient conditions in Ulrich and Watson [1994] for the
construction of monotone quintic spline interpolants of monotone data.
Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
\cdots <X_n$ of breakpoints with corresponding monotone increasing
function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
increasing quintic spline $Q(x)$ with the same breakpoints satisfying
$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
more general, producing $Q(x)$ that is monotone increasing (decreasing)
wherever the data is monotone increasing (decreasing).)

>>>>

The code by Fritsch [1982] for $C^1$ cubic spline interpolation is the
predominantly utilized code for constructing monotone interpolants at
present. Theory has been provided for the quintic case [Ulrich and
Watson 1994; He{\ss} and Schmidt 1994] and that theory was recently
utilized in a proposed algorithm [Lux 2020] for monotone quintic
spline construction, however no published mathematical software
exists. The software presented here represents the first published
software package for producing $C^2$ shape preserving splines. This
work improves upon the algorithm presented by Lux et al. [2020] by
refactoring computations for improved numerical stability, estimating
minimum curvature derivatives at breakpoints with a quadratic facet
model, and using a binary search to reduce the magnitude of the
modifications made to initial derivative estimates when constructing a
monotone spline interpolant.

This work provides a Fortran 2003 subroutine {\tt MQSI} based on the
necessary and sufficient conditions in Ulrich and Watson [1994] for the
construction of monotone quintic spline interpolants of monotone data.
Precisely, the problem is, given a strictly increasing sequence $X_1<X_2<
\cdots <X_n$ of breakpoints with corresponding monotone increasing
function values $Y_1\le Y_2\le \cdots \le Y_n$, find a $C^2$ monotone
increasing quintic spline $Q(x)$ with the same breakpoints satisfying
$Q(X_i)=Y_i$ for $1\le i\le n$. ({\tt MQSI} actually does something slightly
more general, producing $Q(x)$ that is monotone increasing (decreasing)
wherever the data is monotone increasing (decreasing).)

### 2021-02-27 [2020-08]_ACMTOMS_(MQSI)/paper/2-monotone-quintic.tex
Updated paper to responde to coauthor comments. Will use the updated diff with master to finish revising 'response.tex' in upcoming commits.

Fritsch and Carlson [1980] originally proposed the use of central differences
to estimate derivatives, however this often leads to extra and unnecessary
{\it wiggles} in the spline when used to approximate second derivatives.
In an attempt to capture the local shape of the data, this package uses a
facet model from image processing [Haralick and Watson 1981] to estimate
first and second derivatives at breakpoints. Rather than picking a local
linear or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.

>>>>

Fritsch and Carlson [1980] originally proposed the use of central
differences to estimate derivatives, however this often leads to extra
and unnecessary {\it wiggles} in the spline when used to approximate
second derivatives. Modern shape-preserving spline implementations use
a weighted harmonic mean to estimate derivative values at breakpoints
[Moler 2008], however this method also yields approximations with
large {\it wiggle}, or whose second derivative functions have large
$L2$ norm over the domain. In an attempt to capture the local shape of
the data while minimizing wiggle, this package uses a facet model from
image processing [Haralick and Watson 1981] to estimate first and
second derivatives at breakpoints. Rather than picking a local linear
or quadratic fit with minimal residual, this work uses a quadratic
facet model that selects the local quadratic interpolant with minimum
magnitude curvature.
